---
title: 'Report #3'
author: "GROUP 9"
classoption: 12pt
output: bookdown::pdf_document2
bibliography: ../../Reference/bib/References.bib
link-citations: true
biblio_style: "apalike"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
require(ggplot2)
require(ggcorrplot)
require(ggthemes)
require(matlib)
require(bookdown)
require(xtable)
require(moments)
require(gridExtra)
```

# EDA

## Read Data

The dataset have 395 records in total with 5 numerical variables and 28 categorical variables. The numerical variables include age, number of school absences, 1st Period grade, 2nd period grade and final period grade. The categorical variables include sex, address(urban/rural), parents highest qualification, family size, commuting time, study time and so on. The table below shows the first 5 records with 10 variables.

```{r, echo=FALSE}
student=read.table("../../Data/CSV/student-mat.csv",sep=";",header=TRUE)
```

\begin{table}[ht]
\centering
\begin{tabular}{rllrlllrrll}
  \hline
 & school & sex & age & address & famsize & Pstatus & Medu & Fedu & Mjob & Fjob \\ 
  \hline
1 & GP & F &  18 & U & GT3 & A &   4 &   4 & at\_home & teacher \\ 
  2 & GP & F &  17 & U & GT3 & T &   1 &   1 & at\_home & other \\ 
  3 & GP & F &  15 & U & LE3 & T &   1 &   1 & at\_home & other \\ 
  4 & GP & F &  15 & U & GT3 & T &   4 &   2 & health & services \\ 
  5 & GP & F &  16 & U & GT3 & T &   3 &   3 & other & other \\ 
  6 & GP & M &  16 & U & LE3 & T &   4 &   3 & services & other \\ 
   \hline
\end{tabular}
\end{table}



## Summary Table

### Categorical Summary

Three categorical variables:\
School - student's school ('GP' - Gabriel Pereira or 'MS' - Mousinho da Silveira)\
Sex - student's sex ('F' - female or 'M' - male)\
Failures - number of past class failures (n if 1<=n<3, else 4)\

The table \@ref(tab:summary) and Figure&nbsp;\@ref(fig:ct) show the count of each categorical variables. There are more students from Gabriel Pereira school than Mousinho da Silveira school. There are slightly higher number of female students compared to male students. Most of the students never failed the other courses before.

```{r CateSum, echo=FALSE}
student$school <- as.factor(student$school)
student$sex <- as.factor(student$sex)
student$failures <- as.factor(student$failures)
df = data.frame(student$school,student$sex,student$failures)
names(df) = c("School", "Sex", "Failures")
#xtable(summary(df))
```

\begin{table}[ht]
\centering
\caption {Summary of Three Categorical Variables}
\begin{tabular}{lll}
  \hline
 School & Sex & Failures \\ 
  \hline
GP:349   & F:208   & 0:312   \\ 
MS: 46   & M:187   & 1: 50   \\ 
  &  & 2: 17   \\ 
  &  & 3: 16   \\ 
   \hline
\end{tabular} (\#tab:summary)
\end{table}


```{r ct, fig.height=4, fig.cap="Barplot for Categorical Variables", fig.pos="center", echo=FALSE, warning=FALSE}
lapply(names(df), function(col) {
  ggplot(df, aes(.data[[col]], ..count..)) + 
    geom_bar(aes(fill = .data[[col]]), position = "dodge")
}) -> lp
grid.arrange(lp[[1]], lp[[2]], lp[[3]], ncol = 3)
```

### Numerical Summary
Five numerical Variables:\
Age - student's age (from 15 to 22)\
Absences - number of school absences (from 0 to 93)\
G1 - first period grade (from 0 to 20)\
G2 - second period grade (from 0 to 20)\
G3 - final grade (from 0 to 20, output target)\

The table \@ref(tab:numsum) shows five numeric variable statistic summary includes: minimum, first quartile, median, third quartile, and maximum.\
The table \@ref(tab:means) and table \@ref(tab:covariance) show five numeric variable means and covariance.

```{r NumSum, echo=FALSE}
df_num = data.frame(student$age, student$absences, 
                    student$G1, student$G2, student$G3)
names(df_num) <- c("age", "absences", "G1","G2","G3")
#summary(df_num)
```

\begin{table}[ht]
\centering
\caption {Summary of Five numerical Variables}
\begin{tabular}{llllll}
  \hline
 & age & absences & G1 & G2 & G3 \\ 
  \hline
Min. & 15.0 & 0.0   & 3.0 & 0.0 &  0.0  \\
1st Qu. & 16.0 & 0.0   & 8.0 & 9.0 &  8.0  \\ 
Median & 17.0 & 4.0   & 11.0 & 11.0 &  11.0  \\
3rd Qu. & 18.0 & 8.0   & 13.0 & 13.0 &  14.0  \\
Max. & 22.0 & 75.0   & 19.0 & 19.0 &  20.0  \\
   \hline
\end{tabular} (\#tab:numsum)
\end{table}

```{r nummean,echo=FALSE}
df_num.mean<-colMeans(df_num)
```

\begin{table}[ht]
\centering
\caption {Means of Five numerical Variables}
\begin{tabular}{lllll}
  \hline
 age & absences & G1 & G2 & G3 \\ 
  \hline
16.696   & 5.709   & 10.909 & 10.714 &  10.415  \\ 
   \hline
\end{tabular} (\#tab:means)
\end{table}

```{r covmat,echo=FALSE}
#cov(df_num)
```

\begin{table}[ht]
\centering
\caption {Covariance of Five numerical Variables}
\begin{tabular}{llllll}
  \hline
 & age & absences & G1 & G2 & G3 \\ 
  \hline
age & 1.628   & 1.790   & -0.271 & -0.689 &  -0.945  \\
absences & 1.790   & 64.050   & -0.824 & -0.957 &  1.256  \\ 
G1 & -0.271   & -0.824   & 11.017 & 10.639 &  12.188  \\
G2 & -0.689   & -0.957   & 10.639 & 14.149 &  15.594  \\
G3 & -0.945   & 1.256   & 12.188 & 15.594 &  20.990  \\
   \hline
\end{tabular} (\#tab:covariance)
\end{table}

\newpage
Figure \@ref(fig:numplt) shows the count of five numerical variables. Red dashed line as mean and a fitted density in purple.

```{r numplt,fig.height=4, fig.cap="Barplot for Numerical Variables", fig.pos="center", echo=FALSE, warning=FALSE}
lp2 <- list()

lp2[[1]] <- ggplot(data.frame(df_num$age), aes(x=df_num$age)) +
  geom_histogram(aes(y=..density..),
                 bins=nclass.FD(df_num$age),
                 col="black", fill="white") +
  geom_vline(xintercept = df_num.mean[1], col="red",linetype="dashed")+
  geom_density(kernel="epanechnikov", size=.3, col="purple", alpha=.1) +
  xlab("Age") +
  ylab("Count") +
  theme(text=element_text(size=12, 
        family="serif"))

lp2[[2]] <- ggplot(data.frame(df_num$absences), aes(x=df_num$absences)) +
  geom_histogram(aes(y=..density..),
                 bins=nclass.FD(df_num$absences),
                 col="black", fill="white") +
  geom_vline(xintercept = df_num.mean[2], col="red",linetype="dashed")+
  geom_density(kernel="epanechnikov", size=.3, col="purple", alpha=.1) +
  xlab("Absences") +
  ylab("Count") +
  theme(text=element_text(size=12, 
        family="serif"))

lp2[[3]] <- ggplot(data.frame(df_num$G1), aes(x=df_num$G1)) +
  geom_histogram(aes(y=..density..),
                 bins=nclass.FD(df_num$G1),
                 col="black", fill="white") +
  geom_vline(xintercept = df_num.mean[3], col="red",linetype="dashed")+
  geom_density(kernel="epanechnikov", size=.3, col="purple", alpha=.1) +
  xlab("G1") +
  ylab("Count") +
  theme(text=element_text(size=12, 
        family="serif"))

lp2[[4]] <- ggplot(data.frame(df_num$G2), aes(x=df_num$G3)) +
  geom_histogram(aes(y=..density..),
                 bins=nclass.FD(df_num$G3),
                 col="black", fill="white") +
  geom_vline(xintercept = df_num.mean[4], col="red",linetype="dashed")+
  geom_density(kernel="epanechnikov", size=.3, col="purple", alpha=.1) +
  xlab("G2") +
  ylab("Count") +
  theme(text=element_text(size=12, 
        family="serif"))

lp2[[5]] <- ggplot(data.frame(df_num$G3), aes(x=df_num$G3)) +
  geom_histogram(aes(y=..density..),
                 bins=nclass.FD(df_num$G3),
                 col="black", fill="white") +
  geom_vline(xintercept = df_num.mean[5], col="red",linetype="dashed")+
  geom_density(kernel="epanechnikov", size=.3, col="purple", alpha=.1) +
  xlab("G3") +
  ylab("Count") +
  theme(text=element_text(size=12, 
        family="serif"))

grid.arrange(lp2[[1]], lp2[[2]], lp2[[3]],lp2[[4]],lp2[[5]],nrow = 2,ncol = 3)
```

## Skewness and kurtosis

Skewness is a statistical numerical method to measure the asymmetry of the distribution or data set.\
Kurtosis is a numerical method in statistics that measures the sharpness of the peak in the data distribution.\

### Skewness
Table \@ref(tab:skew) shows skewness of five numerical variables.\
Age, absences, and G1 have positive skewness, the graph is said to be positively skewed with the majority of data values less than mean, most of the values are concentrated on the left side.\
Absences skewness is the biggest among all five values, the graph is more concentrated on the left side than age and G1.\
G1 skewness is closest to $0$, the graph is said likely to be symmetric.\
G2 and G3 have negative skewness, the graph is said to be negatively skewed with the majority of data values greater than mean,  most of the values are concentrated on the right side of the graph.\
G3 has a smaller skewness so the graph is more concentrated on the right side than G2.\

```{r skew,echo=FALSE}
#skewness(df_num)
```

\begin{table}[ht]
\centering
\caption {Skewness of Five numerical Variables}
\begin{tabular}{lllll}
  \hline
 age & absences & G1 & G2 & G3 \\ 
  \hline
0.464   & 3.658   & 0.240 & -0.430 &  -0.730  \\ 
   \hline
\end{tabular} (\#tab:skew)
\end{table}

### Kurtosis

Table \@ref(tab:kurt) shows kurtosis of five numerical variables.\
All five values are positive, value of G1 is smaller than $3$, then the data distribution is platykurtic.\
Age is approximately close to $3$, then the data distribtuion is mesokurtic.\
Absences has the biggest kurtosis value, it is much bigger than $3$, then the data distribution is leptokurtic and shows a sharp peak on the graph.\
G2 and G3 are slightly bigger than $3$, then the data distribution is leptokurtic, but the sharp peak on the graph is not so obvious.\

```{r kurt,echo=FALSE}
#kurtosis(df_num)
```

\begin{table}[ht]
\centering
\caption {Kurtosis of Five numerical Variables}
\begin{tabular}{lllll}
  \hline
 age & absences & G1 & G2 & G3 \\ 
  \hline
2.984   & 24.430   & 2.300 & 3.605 &  3.383  \\ 
   \hline
\end{tabular} (\#tab:kurt)
\end{table}

### Visualisation

For both skewness and kurtosis, Figure \@ref(fig:numplt) shows the distribution of five numerical variables, and the sharpness of the peak.

## Correlation Matrix and plot

Table \@ref(tab:correlation) and Figure \@ref(fig:corplt) shows the correlation between all five numerical variables.\
From the table and figure we get that G1, G2 and G3 are have a strong positive correlation.\
Between G2 and G3 have the strongest positive correlation, which is 0.905.\
Between age and G1, G2, G3, they have a low negative correlation.\

```{r cormat,echo=FALSE}
#cor(df_num)
```

\begin{table}[ht]
\centering
\caption {Correlation Matrix of Five numerical Variables}
\begin{tabular}{llllll}
  \hline
 & age & absences & G1 & G2 & G3 \\ 
  \hline
age & 1.000   & 0.175   & -0.064 & -0.143 &  -0.162  \\
absences & 0.175   & 1.000   & -0.031 & -0.032 &  0.034  \\ 
G1 & -0.064   & -0.031   & 1.000 & 0.852 &  0.801  \\
G2 & -0.143   & -0.032   & 0.852 & 1.000 &  0.905  \\
G3 & -0.162   & 0.034   & 0.801 & 0.905 &  1.000  \\
   \hline
\end{tabular} (\#tab:correlation)
\end{table}

```{r corplt,fig.height=4, fig.cap="Correlation between five Numerical Variables", fig.pos="center", echo=FALSE, warning=FALSE}
ggcorrplot(cor(df_num),
method = "circle",
hc.order = TRUE,
type = "lower")
```

## Box Plot

The Figure&nbsp;\@ref(fig:box) shows the box plots of final grade, number of absences and age grouped by the three categorical variables (School, Sex and Failures). The box plots show that the final grade does not differ much among school or sex but it differs by the number of past failures. Even though there are variations, those with fewer number of failures tend to have higher final grades. All box plot for the number of absences shows that they are heavily skewed to large number. The box plot for age shows that Mousinho da Silveira school have higher median age compared to Gabriel Pereira school.



```{r box, fig.height=6, fig.cap="Boxplot of Final grade against 3 categorical variables", echo=FALSE, warning=FALSE, message=FALSE}
lp <- list()
df_all <- cbind(df, df_num)
for (i in 1:length(names(df))){
   lp[[i]] <- ggplot(df_all, aes_string(x=names(df)[i], y="G3", col = names(df)[[i]])) + 
    geom_boxplot(notch=TRUE) +
     labs(y="Final Grade")
}

lp2 <- list()
for (i in 1:length(names(df))){
   lp2[[i]] <- ggplot(df_all, aes_string(x=names(df)[i], y="absences", col = names(df)[[i]])) + 
    geom_boxplot(notch=TRUE) +
     labs(y="Number of Absences")
}

lp3 <- list()
for (i in 1:length(names(df))){
   lp3[[i]] <- ggplot(df_all, aes_string(x=names(df)[i], y="age", col = names(df)[[i]])) + 
    geom_boxplot(notch=TRUE) +
     labs(y="Age")
}


grid.arrange(lp[[1]], lp[[2]], lp[[3]],
             lp2[[1]], lp2[[2]], lp2[[3]],
             lp3[[1]], lp3[[2]], lp3[[3]],ncol = 3)
```

\newpage
# References 
